Staged Abstract Interpreters

### 2019-09-10 paper/20_prelim.tex
polish

The Scala compiler translates the above @for@-comprehension expression into
an equivalent one using @flatMap@ and @map@. The last binding
in @for@-comprehensions is translated into a @map@, where the argument of
@yield@ becomes the body expression inside of that @map@ operation. The
bindings before the last one are all translated into calls of @flatMap@:
\begin{lstlisting}
  xs.flatMap { case x => ys.map { case y => (x, y) } }
\end{lstlisting}

@@

In this section, we present the generic interface in the style of a big-step
definitional interpreter. The key idea is to keep both the binding-time type and
returned monadic type abstract, so that they can be instantiated
differently; we also need to abstract the primitive operations on those types.
In later sections, we will instantiate the monadic type to perform concrete
interpretation \citet{DBLP:conf/popl/LiangHJ95} or abstract interpretation
\cite{Sergey:2013:MAI:2491956.2491979, DBLP:journals/pacmpl/DaraisLNH17}.

\paragraph{Basic Types} We start with some basic type definitions that are used
in the interpreter. The identifiers in the program are represented by strings.
To represent states, two required components for the interpreter are
environments, represented by type @Env@, and stores, represented by type
@Store@. An @Env@ maps identifiers to addresses, and a @Store@ maps addresses to
values. The @Env@ captures bound variables in the scope of current control, and
@Store@ models a persistent heap through the program run-time.  Currently,
the domains of addresses and values are still abstract; hence, they are
declared as abstract types.
\begin{lstlisting}
  trait Semantics {
    type Ident = String; type Addr; type Value
    type Env = Map[Ident, Addr]; type Store = Map[Addr, Value]
    type R[_] // Binding-time as a higher-kinded type
    ... // The definitions in the rest of this section are enclosed in trait Semantics.
  }
\end{lstlisting}

@@

\paragraph{Monadic Operations} We define the return type of the interpreter
@Ans@ as a monadic type @AnsM[_]@ wrapping the type @Value@.
As mentioned in Section~\ref{monadscala}, in order to use the @for@-comprehension
syntax certain operations must be added on the type @AnsM@. Here, we use a
structural type @MonadOps@ to require that type @AnsM@ must (at least)
implement @map@ and @flatMap@. It is worth noting that @MonadOps@ takes another
type parameter @R[_]@ as the binding-time, i.e., @MonadOps[R, AnsM, T]@.
Inside of @MonadOps@, @R[_]@ annotates the data types @A@ and @B@ that are
encapsulated by the monad, but not the monad type @M@ itself. When the generic
interpreter acts as a compiler, we will replace the monads with the ones that
work on staged data values.

@@

\paragraph{The Interpreter} We can now construct the semantics-agnostic interpreter
in monadic form, shown in Figure \ref{fig:shared_int}. The basic idea
of generic interpretation is to traverse the abstract syntax tree
while maintaining the effects (such as environment and state updates).
Note that the interpreter is written in open-recursive style: it cannot
refer to itself directly. Instead, @eval@ takes an additional parameter @ev@ of
type @Expr => Ans@ that refers to itself. Consequently, the method @close@ that
lifts lambda terms to closures and method @ap_clo@ that applies functions also
takes an extra @ev@, because evaluations may happen inside of them.
To close the open recursion, we use a fixed-point operator @fix@.
For concrete-interpretation instantiation, @fix@ works like the Y combinator;
for abstract-interpretation instantiation, it will instrument the interpreter
by memoizing @ev@'s inputs and outputs, which ensures the termination of
abstract interpretation.
Finally, there is a top-level wrapper @run@. The return type @Result@
depends on the kind of monads being used and is therefore also left abstract.
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans
  type Result; def run(e: Expr): Result
\end{lstlisting}

>>>>

The Scala compiler translates the above @for@-comprehension expression into
an equivalent one using @flatMap@ and @map@. The last binding
in @for@-comprehensions is translated into a @map@, where the argument of
@yield@ becomes the body expression inside that @map@ operation. The
bindings before the last one are all translated into calls of @flatMap@:
\begin{lstlisting}
  xs.flatMap { case x => ys.map { case y => (x, y) } }
\end{lstlisting}

@@

In this section, we present the generic interface in the style of a big-step
definitional interpreter. The key idea is to keep both the binding-time type and
returned monadic type abstract, so that they can be instantiated
differently. We also need to abstract the primitive operations on those types.
In later sections, we will instantiate the monadic type to perform concrete
interpretation \citet{DBLP:conf/popl/LiangHJ95} or abstract interpretation
\cite{Sergey:2013:MAI:2491956.2491979, DBLP:journals/pacmpl/DaraisLNH17}.

\paragraph{Basic Types} We start with some basic type definitions that are used
in the interpreter. The identifiers in the program are represented by strings.
To represent states, two required components for the interpreter are
environments, represented by type @Env@, and stores, represented by type
@Store@. An @Env@ maps identifiers to addresses, and a @Store@ maps addresses to
values. An environment captures bound variables in the scope of current control, and
@Store@ models a persistent heap through the program run-time.  Currently,
the domains of addresses and values are still abstract; hence, they are
declared as abstract types.
\begin{lstlisting}
  trait Semantics {
    type Ident = String; type Addr; type Value
    type Env = Map[Ident, Addr]; type Store = Map[Addr, Value]
    type R[_] // Binding-time as a higher-kinded type
    ... // The definitions in the rest of this section are enclosed in trait Semantics.
  }
\end{lstlisting}

@@

\paragraph{Monadic Operations} We define the return type of the interpreter
@Ans@ as a monadic type @AnsM[_]@ wrapping the type @Value@.
As mentioned in Section~\ref{monadscala}, in order to use the @for@-comprehension
syntax certain operations must be added on the type @AnsM@. Here, we use a
structural type @MonadOps@ to require that type @AnsM@ must (at least)
implement @map@ and @flatMap@. It is worth noting that @MonadOps@ takes another
type parameter @R[_]@ as the binding-time, i.e., @MonadOps[R, AnsM, T]@.
Inside @MonadOps@, @R[_]@ annotates the data types @A@ and @B@ that are
encapsulated by the monad, but not the monad type @M@ itself. When the generic
interpreter acts as a compiler, we will replace the monads with the ones that
work on staged data values.

@@

\paragraph{The Interpreter} We can now construct the semantics-agnostic interpreter
in monadic form, shown in Figure \ref{fig:shared_int}. The basic idea
of generic interpretation is to traverse the abstract syntax tree
while maintaining the effects (such as environment and state updates).
Note that the interpreter is written in open-recursive style: it cannot
refer to itself directly. Instead, @eval@ takes an additional parameter @ev@ of
type @Expr => Ans@ that refers to itself. Consequently, the method @close@ that
lifts lambda terms to closures and method @ap_clo@ that applies functions also
takes an extra @ev@, because evaluations may happen inside them.
To close the open recursion, we use a fixed-point operator @fix@.
For concrete-interpretation instantiation, @fix@ works like the Y combinator;
for abstract-interpretation instantiation, it will instrument the interpreter
by memoizing @ev@'s inputs and outputs, which ensures the termination of
abstract interpretation.
Finally, there is a top-level wrapper @run@. The return type @Result@
depends on the kind of monads being used and is therefore also left abstract.
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans
  type Result; def run(e: Expr): Result
\end{lstlisting}
